---
title: "R Basics (part 2)"
author: "Patrick Kimes"
output: html_document
---

## Introduction

We will be covering a few basic _R_ commands in this document as an
introduction to the language and using the RStudio development environment.
This is a [Rmarkdown](https://rmarkdown.rstudio.com) document that includes
both markdown (text) and code chunks. 

## Basic Commands

Performing addition in _R_ is straightforward.

```{r}
3 + 5
```

We can also assign values to variables using `<-`.

```{r}
x1 <- 100
x1
```

To create ranges of integer values, use `:`.

```{r}
x2 <- 1:7
x2
```

Alternatively, we can call the `seq()` function to create a sequence of
values between a start and end value, with values offset by a specified
value.

```{r}
xSkip <- seq(1, 7, by = 2)
xSkip
```

Most _R_ functions come with great (or at least some) documentation that
can be easily accessed from the console. For example, if we need some
help with the `seq()` function from above, we can call either of the
following commands.

```{r, eval = FALSE}
?seq
help(seq)
```

Notice that `xSkip` is a vector of values. To access a subset of the range,
use the `[A:B]` notation. 

```{r}
xSkip[2:3]
```

Unlike many other programming languages, indexing in _R_ is 1-based
(and not 0-based). This means that the first entry in the vector is
`b[1]` and not `b[0]` (which is undefined).

Vectors can be constructed using `c()`. 

```{r}
xvec <- c(2, 9, 1, 8, 2)
xvec

yvec <- c("a", "b", "c", "d", "e")
yvec
```

All elements in vector must be of the same type. Let's see what
happens when we try to create a vector with both numbers and letters.

```{r}
c("a", "b", 3, 4, "e")
```

_R_ converted our numbers to characters!

Aside from numbers and characters, the _R_ language also includes a
special **factor** type. Factors are ordered vector of "labels" that
(for statisticians) intuitively translate to levels in an experimental
design.

```{r}
fvec <- factor(c("apple", "grape", "apple", "apple"),
               levels = c("grape", "apple"))
fvec
```

While factors look like character strings, **they're not**! This is a
common pitfall when starting off in _R_. We can pull out the factor
levels with the `levels()` function.

```{r}
levels(fvec)
```

If we have factors but need characters, we can change the type
by calling `as.character()`.

```{r}
as.character(fvec)
```

Finally, let's talk about _data.frames_, how tabular data is stored
and organized in _R_. Start by creating a new data.frame with two
columns, `someLetters` and `someNumbers` with the following code.

```{r}
df <- data.frame(someLetters = yvec,
                 someNumbers = xvec)
df
```

There are a couple different ways to access columns in the data.frame.

```{r}
df$someLetters
df[, "someLetters"]
df[, 1]
```

Interesting! Notice how our column is a factor. This is a little surprising
since `yvec` was a vector of characters.

For historical reasons, by default, character columns in data.frames are
turned into factors in _R_. We can disable this when creating the data.frame
by specifying the `stringsAsFactors=` parameter.

```{r}
df <- data.frame(someLetters = yvec,
                 someNumbers = xvec,
                 stringsAsFactors = FALSE)
df$someLetters
```

Great! Now we're comfortable with some of the major data types used in _R_
for data science. (We've skipped matrices in the interest of time,
but make sure to read about these when you get a chance!)

Here are a few other useful commands when getting started with _R_.

To see what variables we have defined in our _R_ session, call `ls()`.

```{r}
ls()
```

The following command is used to install a new package
(**this only needs to be done once and when updating**).

```{r, eval = FALSE}
install.packages("NewPackageIWant")
```

Whenever we open up _R_ in a new _session_ and need a package,
we will need to load the package with `library()`.

```{r, eval = FALSE}
library("NewPackageIWant")
```


## Example Analysis

We will now walk through an example analysis using a dataset that
we've included in the `data` folder as a `CSV` file.

Let's read in the data.

```{r}
## use read.csv() to read data and store as 'dat'
dat <- read.csv("../data/example-data.csv")
```

Next, let's take a look at the data.

```{r}
## call the object to print out the content
dat
```

Let's go over a few more basic commands to see what the data
looks like.

```{r}
## use head() to see top rows of table
head(dat)

## use dim() to check dimensions of table
dim(dat)
```

Interesting! Let's take a closer look at the `type` column.

```{r}
## use $ to pull out the 'type' column
dat$type

## use table() to tabulate the 'type' column
table(dat$type)
```

It looks like we have a balanced set of observations from
two groups. Let's take a look at the `measurement` column.

```{r}
## use summary() to summarize the 'measurement' column
summary(dat$measurement)
```

Let's create a boxplot of the data for the two groups.

```{r}
## use boxplot() to create a boxplot
boxplot(measurement ~ type, data = dat)
```

Boxplots are a great visualization tool, but they can sometimes
summarize the data too much. Let's actually plot the data.

```{r}
## use boxplot() and points() to create a boxplot
boxplot(measurement ~ type, data = dat)
points(measurement ~ type, data = dat)
```

It looks like we have outliers! Let's drop them.

```{r}
## remove outliers using [, ] to subset rows
dat <- dat[dat$measurement < 9999, ]
```

```{r}
## use boxplot() and points() to create a boxplot
boxplot(measurement ~ type, data = dat)
points(measurement ~ type, data = dat)
```

It looks like the data might be heavy-tailed.
We can also take a look at the data after log2-transforming
the rows.

```{r}
## use log2 to transform the 'measurement' column and replot
boxplot(log2(measurement) ~ type, data = dat)
points(log2(measurement) ~ type, data = dat)
```

Depending on what `measurement` represents, it may be better
to move forward with our analysis using the log2-transformed
values.

```{r}
## use $ and <- to replace the `measurement` column
dat$measurement <- log2(dat$measurement)
dat
```

Be careful not to run the above code more than once!

Finally, we can run a simple t-test with the `t.test()`.

```{r}
## use t.test to perform a two sample t-test
t.test(measurement ~ type, data = dat)
```

When we're done, try `knit`ing this document to see what you get!
